---
title: "Qualitative Lesson Analysis Pipeline"
author: "WB"
date: "2025-11-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Qualitative Lesson Analysis Pipeline

This pipeline performs qualitative analysis of Learnable lessons using OpenAI's API.

## 1. Load Dependencies and Global Setup

```{r dependencies}
# Source global setup (database connection, caching, text extraction functions)
source("get_analysis_global.R")

# Additional packages for API calls
library(httr)
library(jsonlite)
library(dotenv)
library(purrr)
library(glue)

# Load environment variables for API key
load_dot_env()
```

## 2. Configuration

```{r config}
# OpenAI API configuration
get_openai_key <- function() {
  api_key <- Sys.getenv("OPENAI_API_KEY")
  if (api_key == "") {
    stop("OPENAI_API_KEY not found in environment variables")
  }
  api_key
}

# Available models for analysis
AVAILABLE_MODELS <- c(
  "gpt-4o",
  "gpt-4.1",
  "gpt-4o-mini",
  "gpt-5.1",                    # GPT-5.1 (released Nov 2025)
  "gpt-5.1-2025-11-13",         # Versioned identifier
  "gpt-5.1-chat-latest"         # GPT-5.1 Instant
)

# Default model for qualitative analysis
DEFAULT_MODEL <- "gpt-5.1"

# API endpoint configuration
# All models use the standard chat completions endpoint
OPENAI_API_URL <- "https://api.openai.com/v1/chat/completions"

get_api_endpoint <- function(model) {
  OPENAI_API_URL
}

# Check if model is GPT-5.1 (supports adaptive reasoning)
is_gpt5_model <- function(model) {
  grepl("^gpt-5", model, ignore.case = TRUE)
}



# Load instructions
INSTRUCTIONS_FILE <- "lesson_analyser_instructions.txt"
SCHEMA_FILE <- "qa_schema.json"
```

## 3. Load Analysis Instructions and Schema

```{r load_instructions}
# Load the qualitative analysis instructions
load_analysis_instructions <- function(file_path = INSTRUCTIONS_FILE) {
  if (!file.exists(file_path)) {
    stop(glue("Instructions file not found: {file_path}"))
  }

  instructions <- paste(readLines(file_path, warn = FALSE), collapse = "\n")
  message("✅ Loaded analysis instructions (", nchar(instructions), " chars)")
  instructions
}

# Load the JSON schema for validation
load_analysis_schema <- function(file_path = SCHEMA_FILE) {
  if (!file.exists(file_path)) {
    stop(glue("Schema file not found: {file_path}"))
  }

  schema <- fromJSON(file_path, simplifyVector = FALSE)
  message("✅ Loaded analysis schema")
  schema
}

qa_instructions <- load_analysis_instructions()
qa_schema <- load_analysis_schema()
```

## 4. API Communication Functions

```{r api_functions}
# Send request to OpenAI API with retry logic
send_openai_request <- function(
  api_key,
  messages,
  model = DEFAULT_MODEL,
  max_tokens = 4000,
  temperature = 1,
  retries = 3,
  reasoning_effort = "medium"  # For GPT-5.1: "low", "medium", "high", or "none"
) {

  if (!is.character(api_key) || nchar(api_key) < 20) {
    stop("Invalid API key")
  }

  if (!is.list(messages) || length(messages) == 0) {
    stop("Messages must be a non-empty list")
  }

  message("  → Sending request to OpenAI (model: ", model, ")")

  # Build request body based on model type
  request_body <- list(
    model = model,
    messages = messages
  )

  # GPT-5.1 uses max_completion_tokens and supports reasoning_effort
  if (is_gpt5_model(model)) {
    request_body$max_completion_tokens <- max_tokens
    request_body$temperature <- temperature
    request_body$response_format <- list(type = "json_object")
    # Add reasoning_effort for adaptive reasoning control
    if (!is.null(reasoning_effort) && reasoning_effort != "") {
      request_body$reasoning_effort <- reasoning_effort
      message("  ℹ️ GPT-5.1 adaptive reasoning: ", reasoning_effort)
    }
  } else {
    # Standard GPT-4 models
    request_body$max_tokens <- max_tokens
    request_body$temperature <- temperature
    request_body$response_format <- list(type = "json_object")
  }

  for (attempt in 1:retries) {
    response <- tryCatch({
      POST(
        url = get_api_endpoint(model),
        add_headers(
          Authorization = paste("Bearer", api_key),
          `Content-Type` = "application/json"
        ),
        encode = "json",
        body = request_body,
        timeout(120)
      )
    }, error = function(e) {
      message("❌ Request error: ", e$message)
      NULL
    })

    if (is.null(response)) {
      if (attempt < retries) {
        wait_time <- 2^attempt
        message("⏳ Retrying in ", wait_time, " seconds...")
        Sys.sleep(wait_time)
        next
      } else {
        stop("Failed to connect to OpenAI API after ", retries, " attempts")
      }
    }

    status <- status_code(response)

    if (status == 200) {
      parsed <- fromJSON(content(response, as = "text", encoding = "UTF-8"), flatten = TRUE)
      message("✅ Request successful")
      return(parsed)
    } else if (status == 429) {
      wait_time <- 2^attempt
      message("⚠️ Rate limit hit. Retrying in ", wait_time, " seconds...")
      Sys.sleep(wait_time)
    } else {
      error_content <- content(response, as = "text", encoding = "UTF-8")
      error_msg <- tryCatch({
        error_json <- fromJSON(error_content)
        if (!is.null(error_json$error$message)) {
          error_json$error$message
        } else {
          error_content
        }
      }, error = function(e) error_content)

      message("❌ API error (status ", status, ")")
      message("   Error details: ", error_msg)

      # For 400 errors, show the request body for debugging
      if (status == 400) {
        message("   Request model: ", request_body$model)
        message("   Request had ", length(request_body$messages), " messages")
      }

      stop("API request failed with status ", status, ": ", error_msg)
    }
  }

  stop("Max retries reached")
}

# Extract and validate the assistant's response
extract_analysis_response <- function(api_response) {
  if (is.null(api_response$choices) || length(api_response$choices) == 0) {
    stop("No choices in API response")
  }

  # Extract the message content
  content <- api_response$choices$message.content[1]

  if (is.null(content) || content == "") {
    stop("Empty content in API response")
  }

  # Parse JSON response
  analysis <- tryCatch({
    fromJSON(content, simplifyVector = FALSE)
  }, error = function(e) {
    message("❌ Failed to parse JSON response: ", e$message)
    message("Raw content: ", substr(content, 1, 500))
    NULL
  })

  if (is.null(analysis)) {
    return(NULL)
  }

  # Extract token usage
  tokens <- list(
    prompt = api_response$usage$prompt_tokens %||% NA,
    completion = api_response$usage$completion_tokens %||% NA,
    total = api_response$usage$total_tokens %||% NA
  )

  list(
    analysis = analysis,
    tokens = tokens,
    finish_reason = api_response$choices$finish_reason[1]
  )
}

# Validate analysis against schema (basic validation - for backward compatibility)
validate_analysis_structure <- function(analysis, schema) {
  result <- validate_analysis_with_feedback(analysis, schema)
  return(result$is_valid)
}

# Enhanced validation with detailed feedback for correction loop
# Now dynamically parses the schema instead of using hardcoded validation
validate_analysis_with_feedback <- function(analysis, schema) {
  errors <- character()

  # Helper: recursively validate an object against a schema node
  validate_node <- function(data, schema_node, path = "") {
    node_errors <- character()

    # Get the schema properties - handle both direct properties and nested type definitions
    schema_props <- schema_node$properties
    if (is.null(schema_props) && !is.null(schema_node$type)) {
      # This might be a type reference, not a properties object
      return(node_errors)
    }

    # Check for required fields if specified in schema
    if (!is.null(schema_node$required)) {
      required_fields <- unlist(schema_node$required)
      missing_fields <- setdiff(required_fields, names(data))

      if (length(missing_fields) > 0) {
        field_path <- if (nchar(path) > 0) paste0(path, ".") else ""
        node_errors <- c(node_errors,
          paste0("Missing required field(s) at '", field_path, "': ",
                 paste(missing_fields, collapse = ", ")))
      }
    }

    # Validate each property that exists in the schema
    if (!is.null(schema_props)) {
      for (prop_name in names(schema_props)) {
        prop_schema <- schema_props[[prop_name]]
        field_path <- if (nchar(path) > 0) paste0(path, ".", prop_name) else prop_name

        # Skip if this field isn't in the data
        if (!prop_name %in% names(data)) {
          next
        }

        prop_data <- data[[prop_name]]

        # Validate type
        expected_type <- prop_schema$type
        if (!is.null(expected_type)) {
          # Handle array types (can be multiple types like ["string", "null"])
          if (is.list(expected_type)) {
            expected_type <- expected_type[[1]]  # Use first type
          }

          valid_type <- FALSE
          if (expected_type == "object") {
            valid_type <- is.list(prop_data) && !is.null(names(prop_data))
          } else if (expected_type == "array") {
            valid_type <- is.list(prop_data)
          } else if (expected_type == "string") {
            valid_type <- is.character(prop_data)
          } else if (expected_type == "integer") {
            valid_type <- is.numeric(prop_data) && prop_data == round(prop_data)
          } else if (expected_type == "number") {
            valid_type <- is.numeric(prop_data)
          } else if (expected_type == "boolean") {
            valid_type <- is.logical(prop_data)
          }

          if (!valid_type) {
            node_errors <- c(node_errors,
              paste0("Field '", field_path, "' has incorrect type. Expected: ",
                     expected_type, ", Got: ", class(prop_data)[1]))
          }
        }

        # Validate integer range (minimum/maximum)
        if (!is.null(prop_schema$minimum) && is.numeric(prop_data)) {
          if (prop_data < prop_schema$minimum) {
            node_errors <- c(node_errors,
              paste0("Field '", field_path, "' value ", prop_data,
                     " is below minimum ", prop_schema$minimum))
          }
        }
        if (!is.null(prop_schema$maximum) && is.numeric(prop_data)) {
          if (prop_data > prop_schema$maximum) {
            node_errors <- c(node_errors,
              paste0("Field '", field_path, "' value ", prop_data,
                     " exceeds maximum ", prop_schema$maximum))
          }
        }

        # Validate enum values
        if (!is.null(prop_schema$enum)) {
          valid_values <- unlist(prop_schema$enum)
          # Filter out NULL from valid values for comparison
          valid_values_non_null <- valid_values[!sapply(valid_values, is.null)]
          if (!is.null(prop_data) && !prop_data %in% valid_values_non_null) {
            node_errors <- c(node_errors,
              paste0("Field '", field_path, "' has invalid value '", prop_data,
                     "'. Must be one of: ", paste(valid_values_non_null, collapse = ", ")))
          }
        }

        # Recurse into nested objects
        if (expected_type == "object" && is.list(prop_data) && !is.null(prop_schema$properties)) {
          nested_errors <- validate_node(prop_data, prop_schema, field_path)
          node_errors <- c(node_errors, nested_errors)
        }

        # Validate array items
        if (expected_type == "array" && is.list(prop_data) && !is.null(prop_schema$items)) {
          for (i in seq_along(prop_data)) {
            item_path <- paste0(field_path, "[", i, "]")
            if (!is.null(prop_schema$items$properties)) {
              item_errors <- validate_node(prop_data[[i]], prop_schema$items, item_path)
              node_errors <- c(node_errors, item_errors)
            }
          }
        }
      }
    }

    return(node_errors)
  }

  # Run validation
  errors <- validate_node(analysis, schema, "")

  # Build feedback message
  if (length(errors) > 0) {
    feedback_message <- paste0(
      "The JSON response has the following validation errors:\n\n",
      paste(paste0(seq_along(errors), ". ", errors), collapse = "\n"),
      "\n\n**Instructions for correction:**\n",
      "- Ensure all field names use snake_case (e.g., 'lesson_typing', not 'Lesson Typing')\n",
      "- All rubric dimensions must be objects with 'score' (integer 0-3) and 'reason' (string) fields\n",
      "- Check that all required fields are present\n",
      "- Verify that numeric values are within the specified ranges\n",
      "- Ensure enum fields contain only valid values\n\n",
      "Please correct these issues and return a complete, valid JSON object that exactly matches the schema."
    )

    return(list(
      is_valid = FALSE,
      errors = errors,
      feedback_message = feedback_message
    ))
  }

  message("✅ Analysis structure validated")
  return(list(is_valid = TRUE, errors = NULL, feedback_message = NULL))
}
```

## 5. Main Analysis Function with Validation Loop

```{r main_analysis}
# Analyze a single lesson using OpenAI API with validation loop
analyze_lesson_qualitative <- function(
  lesson_text,
  lesson_id = NULL,
  lesson_title = NULL,
  model = DEFAULT_MODEL,
  api_key = NULL,
  include_metadata = TRUE,
  max_iterations = 3,
  reasoning_effort = "medium"  # For GPT-5.1: "low", "medium", "high", or "none"
) {

  start_time <- Sys.time()

  # Get API key
  if (is.null(api_key)) {
    api_key <- get_openai_key()
  }

  # Validate inputs
  if (is.null(lesson_text) || nchar(lesson_text) < 100) {
    stop("Lesson text is too short or missing")
  }

  # Estimate token usage
  estimated_tokens <- nchar(lesson_text) / 4 + nchar(qa_instructions) / 4
  message("📊 Estimated input tokens: ~", round(estimated_tokens))

  if (estimated_tokens > 25000) {
    warning("⚠️ Lesson text is very long and may exceed model limits")
  }

  # Build metadata preamble if requested
  metadata_text <- ""
  if (include_metadata && (!is.null(lesson_id) || !is.null(lesson_title))) {
    metadata_parts <- c()
    if (!is.null(lesson_id)) {
      metadata_parts <- c(metadata_parts, paste("Lesson ID:", lesson_id))
    }
    if (!is.null(lesson_title)) {
      metadata_parts <- c(metadata_parts, paste("Lesson Title:", lesson_title))
    }
    metadata_text <- paste(
      "\n\n--- Lesson Metadata ---\n",
      paste(metadata_parts, collapse = "\n"),
      "\n--- End Metadata ---\n\n",
      sep = ""
    )
  }

  # Construct the user prompt
  user_prompt <- paste0(
    metadata_text,
    "--- Lesson Content ---\n\n",
    lesson_text,
    "\n\n--- End Lesson Content ---\n\n",
    "Please analyze this lesson according to the instructions and return a JSON object matching the schema provided in the system instructions."
  )

  # Build initial message list
  messages <- list(
    list(role = "system", content = qa_instructions),
    list(role = "user", content = user_prompt)
  )

  # Initialize token counters
  total_token_count <- 0
  prompt_token_count <- 0
  completion_token_count <- 0

  # Validation loop (similar to airlab chat_loop_stepwise)
  message("🚀 Starting qualitative analysis with validation loop...")

  for (iteration in 1:max_iterations) {
    message("🔄 Iteration ", iteration, "/", max_iterations)

    # Send request
    api_response <- send_openai_request(
      api_key = api_key,
      messages = messages,
      model = model,
      max_tokens = 4000,
      temperature = 1,
      reasoning_effort = reasoning_effort
    )

    # Extract analysis
    extracted <- extract_analysis_response(api_response)

    if (is.null(extracted)) {
      stop("Failed to extract analysis from API response")
    }

    # Accumulate token counts
    total_token_count <- total_token_count + extracted$tokens$total
    prompt_token_count <- prompt_token_count + extracted$tokens$prompt
    completion_token_count <- completion_token_count + extracted$tokens$completion

    # Validate structure
    validation_result <- validate_analysis_with_feedback(extracted$analysis, qa_schema)

    if (validation_result$is_valid) {
      # Success - return result
      elapsed_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

      result <- list(
        lesson_id = lesson_id,
        lesson_title = lesson_title,
        model = model,
        timestamp = Sys.time(),
        elapsed_time = elapsed_time,
        tokens = list(
          prompt = prompt_token_count,
          completion = completion_token_count,
          total = total_token_count
        ),
        analysis = extracted$analysis,
        validation_passed = TRUE,
        iterations = iteration
      )

      message("✅ Analysis complete in ", round(elapsed_time, 1), " seconds")
      message("📊 Total tokens used: ", total_token_count,
              " (prompt: ", prompt_token_count,
              ", completion: ", completion_token_count, ")")
      message("🔁 Converged after ", iteration, " iteration(s)")

      return(result)
    } else {
      # Validation failed - add feedback to conversation
      message("⚠️ Validation failed on iteration ", iteration)
      message("   Errors: ", paste(validation_result$errors, collapse = "; "))

      # Add assistant's response and user feedback to conversation
      messages <- append(messages, list(
        list(role = "assistant", content = toJSON(extracted$analysis, auto_unbox = TRUE)),
        list(role = "user", content = validation_result$feedback_message)
      ))

      # Continue to next iteration
      if (iteration < max_iterations) {
        message("   Retrying with feedback...")
        Sys.sleep(1)  # Brief pause
      }
    }
  }

  # If we get here, validation failed after all iterations
  elapsed_time <- as.numeric(difftime(Sys.time(), start_time, units = "secs"))

  warning("❌ Analysis failed validation after ", max_iterations, " iterations")

  result <- list(
    lesson_id = lesson_id,
    lesson_title = lesson_title,
    model = model,
    timestamp = Sys.time(),
    elapsed_time = elapsed_time,
    tokens = list(
      prompt = prompt_token_count,
      completion = completion_token_count,
      total = total_token_count
    ),
    analysis = extracted$analysis,
    validation_passed = FALSE,
    iterations = max_iterations,
    final_errors = validation_result$errors
  )

  return(result)
}
```

## 6. Save and Load Results

```{r save_load}
# Save a single analysis result with pretty formatting for easy inspection
save_analysis_result <- function(result, output_dir = "qualitative_analysis_results") {
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
    message("📁 Created output directory: ", output_dir)
  }

  # Create filename with lesson ID and timestamp
  filename <- paste0(
    "lesson_",
    result$lesson_id,
    "_",
    format(result$timestamp, "%Y%m%d_%H%M%S"),
    ".json"
  )

  filepath <- file.path(output_dir, filename)

  # Write JSON with pretty formatting for easy inspection
  write_json(
    result,
    filepath,
    pretty = TRUE,
    auto_unbox = TRUE,
    null = "null"
  )

  message("💾 Saved analysis to: ", filepath)
  invisible(filepath)
}

# Load a saved analysis result
load_analysis_result <- function(filepath) {
  if (!file.exists(filepath)) {
    stop("File not found: ", filepath)
  }

  result <- fromJSON(filepath, simplifyVector = FALSE)
  message("✅ Loaded analysis from: ", filepath)
  result
}

# Helper function to view just the analysis portion (without metadata)
view_analysis <- function(result) {
  result$analysis
}

# Helper function to create a human-readable summary of a result
summarize_result <- function(result) {
  analysis <- result$analysis

  cat("=== LESSON ANALYSIS SUMMARY ===\n\n")
  cat("Lesson ID:", result$lesson_id, "\n")
  cat("Lesson Title:", result$lesson_title %||% "N/A", "\n")
  cat("Model:", result$model, "\n")
  cat("Timestamp:", format(result$timestamp), "\n")
  cat("Elapsed Time:", round(result$elapsed_time, 1), "seconds\n")
  cat("Tokens Used:", result$tokens$total,
      "(prompt:", result$tokens$prompt, "+ completion:", result$tokens$completion, ")\n")
  cat("Validation Passed:", result$validation_passed, "\n")
  cat("Iterations:", result$iterations, "\n\n")

  cat("--- LESSON TYPING ---\n")
  cat("Stated Type:", analysis$lesson_typing$stated_type$stage %||% "N/A",
      "/", analysis$lesson_typing$stated_type$sub_type %||% "N/A", "\n")
  cat("Inferred Type:", analysis$lesson_typing$inferred_type$stage %||% "N/A",
      "/", analysis$lesson_typing$inferred_type$sub_type %||% "N/A", "\n\n")

  cat("--- ATOMIC LESSON ASSESSMENT ---\n")
  cat("Is Single Atomic Lesson:", analysis$atomic_lesson_assessment$is_single_atomic_lesson, "\n")
  cat("Number of Arcs:", length(analysis$atomic_lesson_assessment$arcs), "\n\n")

  cat("--- RUBRIC SCORES (0-3) ---\n")
  rubric <- analysis$rubric
  cat("Template Fit:", rubric$template_fit$score, "\n")
  cat("Structure Compliance:", rubric$template_structure_compliance$score, "\n")
  cat("Cognitive Integrity:", rubric$cognitive_integrity$score, "\n")
  cat("Boundary Discipline:", rubric$template_boundary_discipline$score, "\n")
  cat("Progressive Model:", rubric$progressive_model_principle$score, "\n")
  cat("Lesson Economy:", rubric$lesson_economy_and_cognitive_load$score, "\n\n")

  cat("--- PRIORITY INDEX ---\n")
  pi <- analysis$priority_index
  cat("Priority Index:", pi$priority_index, "\n")
  cat("Priority Band:", pi$priority_band, "\n")
  cat("Base Sum:", pi$base_sum, "\n")
  cat("Overload Factor:", pi$overload_factor, "\n\n")
}
```

## 7. Example Usage

```{r example_usage, eval=FALSE}
# --- Single Lesson Analysis ---

# Select a lesson ID to analyze
example_lesson_id <- "2311"

# Get lesson text (using functions from get_analysis_global.R)
example_lesson_text <- get_updated_lesson_text(
  selected_lesson_ids = example_lesson_id,
  cached_lesson_text = cached_lesson_text,
  content_tree = content_tree
)

# Run analysis
single_result <- analyze_lesson_qualitative(
  lesson_text = example_lesson_text$text[1],
  lesson_id = example_lesson_id,
  lesson_title = example_lesson_text$lesson[1]
)

# Save the result
save_analysis_result(single_result)

# View human-readable summary
summarize_result(single_result)


```

```{r}

# View just the analysis portion
view_analysis(single_result)

# View priority index details
#single_result$analysis$priority_index

# View rubric scores with reasons
#single_result$analysis$rubric$template_fit
```




```{r}
# Example: Analyze and inspect a lesson
single_result
```


```{r}

```

---

## Notes

- This pipeline is designed as a black box function to be called from get_analysis_global.R
- The main function is `analyze_lesson_qualitative()` which takes lesson text and returns structured analysis
- Results are automatically validated against the schema with iterative feedback correction
- Save results with `save_analysis_result()` for easy JSON inspection
- Use `summarize_result()` to quickly view key metrics without opening the JSON file

## Integration with get_analysis_global.R

To use this pipeline from get_analysis_global.R, iterate through lessons and call:

```r
# In get_analysis_global.R:
for (lesson_id in selected_lesson_ids) {
  lesson_data <- get_updated_lesson_text(lesson_id, cached_lesson_text, content_tree)

  result <- analyze_lesson_qualitative(
    lesson_text = lesson_data$text[1],
    lesson_id = lesson_id,
    lesson_title = lesson_data$lesson[1]
  )

  save_analysis_result(result, output_dir = "qualitative_analysis_results")
}
```

---

## Removed Sections

The following sections were removed as batch processing should be handled in get_analysis_global.R:
- `analyze_lessons_batch()` function
- `summarize_priority_indices()`, `summarize_rubric_scores()`, `summarize_lesson_typing()` functions

These summary functions can be recreated in get_analysis_global.R by loading multiple saved JSON files if needed.
